{\TheTitle}\thanks{Submitted to the editors DATE. \funding{This work was funded by the Fog Research Institute under contract no.~FRI-454.}}

### 2017-10-19 ssc-0-0.tex
revise the semi-implicit paper into a second draft

% formula equations
The semi-implicit formula is a rule for integrating systems of
ordinary differential equations \cite{jkp-132-312,jkp-93-144}.  It
is often used as a method of lines, to solve stochastic partial
differential equations that occur when phase space Monte-Carlo
methods are used in quantum optics and in atom optics \cite{pra-58-4824}.
These equations are related to Schr\"odinger's equation,
\begin{equation}\label{eq:vi}
    \frac{\partial\psi}{\partial t} = -i(-\nabla^2 + V)\psi,
\end{equation}
with additional nonlinear and random terms.  The spatial axes are
discretised, often using spectral derivative operators, to give a
system of ordinary or stochastic differential equations
\cite{2000-Trefethen-Spectral}.  Then these equations are integrated
by the semi-implicit formula.

@@

% basis coefficients
When the semi-implicit rule is applied to \cref{eq:vi}, or a
stochastic or nonlinear variant of it, the laplacian operator is usually
evaluated by a spectral method \cite{something}.  This is equivalent to expanding
the wave function~$\psi(x)$ over a basis of sinc wave packets
\begin{equation}
	\psi(x) = \sum_m s^m\phi_m(x),
\end{equation}
where, for a spatial grid of step~$l$,
\begin{equation}
	\phi_m(x) = \frac1{\sqrt{l}} \sinc\left(\textstyle\frac{x}{l} - m\right).
\end{equation}
This paper uses the convention
\begin{equation}
	\sinc x = \frac{\sin\pi x}{\pi x},
\end{equation}
under which the set of functions~$\phi_m$ is an orthonormal basis for the
space of functions limited to spatial bandwidth~$1/2l$.  The spectral method
reduces \cref{eq:vi} to an equation for the complex coefficient vector~$s$ of the form
\begin{equation}\label{eq:iii}
	{ds\over dt}=-iHs,
\end{equation}
where~$H$ is a Hermitian matrix with elements
\begin{equation}
	H_{mn} = \int \phi_m(x) (-\nabla^2 + V(x)) \phi_n^\ast(x)\,dx.
\end{equation}

@@

Suppose that the point~$s_{n+1}$ also lies on this circle.
Then~$f_{n+\frac12}=-\frac{i}2H(s_n+s_{n+1})$ is parallel to the
chord from~$s_n$ to~$s_{n+1}$.  The midpoint of the chord can be
brought towards and away from the origin by moving~$s_{n+1}$ around
the circle, so there is some position for
which~$|s_{n+1}-s_n|=h|f_{n+\frac12}|$, and this value of~$s_{n+1}$
satisfies Equation~\cref{eq:i}.  The only error introduced by the
formula is that~$c_k$ rotates with an angular frequency less
than~$\omega_k$, which is bounded by~$\pi/h$ as~$\omega_k$ becomes
large.

In practice, the iterations of~Equation~\cref{eq:vi} must be used
instead of~Equation~\cref{eq:i}, and this is less stable.  For
Hamiltonian dynamics with the form of Equation~\cref{eq:iii}, the
iteration rule becomes
\begin{equation}
s_{n+1}^{(j+1)}=s_n+hf_{n+\frac12}^{(j)}\qquad f_{n+\frac12}^{(j+1)}
	=-{i\over2}Hs_n-{i\over2}Hs_{n+1}^{(j)}
\end{equation}
or
\begin{equation}
s_{n+1}^{(j+1)}=\left(1-{ih\over2}H\right)s_n-{ih\over2}Hs_{n+1}^{(j)}
\end{equation}
with solution
\begin{equation}\label{eq:viii}
s_{n+1}^{(j)}=U^{(j)}s_n
	=\left(1+2\sum_{i=1}^j\left({-ihH\over 2}\right)^j\right)s_n.
\end{equation}
The time step matrix~$U^{(j)}$ for~$j$ iterations has the the same
eigenvectors~$v_k$ as the hamiltonian matrix~$H$.  Each of its
eigenvalues~$u_k^{(j)}$ is the sum of a geometric series
in~$-ih\omega_k/2$, which agrees to second order in~$\omega_k$ with
the eigenvalue~$e^{-ih\omega_k}$ of the exact time step
operator~$U=e^{-ihH}$.  The relation between~$h\omega_k$ and~$u_k^{(j)}$
plotted in \cref{fig:ii}, for the first four iterations.

\begin{figure}[htbp]
  \centering
  \includegraphics[width=\textwidth]{wev.eps}
  \caption{Relation between eigenvectors~$\omega_k$ of~$H$ and~$u_k^{(j)}$ of~$U^{(j)}$.  The lightest line is for~$j=1$, and the heavier ones are for~$j=2$, 3, 4.  The exact time evolution operator~$U$ has unimodular eigenvalues.}
  \label{fig:ii}
\end{figure}

The geometric series for~$U^{(j)}$ will converge if and only if the
angular frequencies~$\omega_k$ are bounded by the reciprocal time
step~$\frac2h$, a kind of Nyquist condition.  If the discretised
hamiltonian has an eigenvalue~$\omega_k\gg \frac2h$, then~$U^{(j)}$
will enlarge the component of~$v_k$ in~$s_n$ by a factor close
to~$2(h\omega_k/2)^j$ over the time step~$h$.  This component is
thus a parasitic solution growing at the
rate~$\frac{\ln2}h+\frac{j}{h}\ln\left(\frac{h\omega_k}2\right)$.

{\bf Rewrite for clarity.}
The semi-implicit method is rarely used to solve the linear
Equation~\cref{eq:iii}, and more often used to solve a nonlinear
Gross-Pitaevskii~equation of the form
\begin{equation}
	{\partial\psi\over\partial t}=-i(\hat H+g|\psi|^2)\psi.
\end{equation}
However, following the usual approach to the stability of integation formulae \cite{something}, the mapping
$\psi\to -i(\hat H+g|\psi|^2)\psi$ can be linearised at each time step, in the form
\begin{equation}
	\psi(t)+\delta\psi\to -i(\hat H+g|\psi(t)|^2)\psi(t) -i\hat H'\delta\psi.
\end{equation}
This linearisation has been extensively studied, and the operator~$\hat
H'$ is the hamiltonian for Bogoliubov sound waves \cite{aop-70-67}.
The stability conditon for the Gross-Pitaevskii~equation is that
the highest frequency of the Bogoliubov sound waves in the discretised
Hamiltonian is less than~$2/h$.  In the limit of fine grids, the
high-frequency Bogoliubov sound waves have the same frequencies as
the eigenstates of~$\hat H$.  Most simulations use grids fine enough
for this to be the case, and the nonlinear term will not affect the
stability of such simulations.

@@

As a test of this analysis, a quantum Kerr oscillator was simulated
using the semi-implicit method.  This system is a single-orbital
analog of the fields for repulsive particles that are simulated in
quantum optics.  The states with definite numbers of particles are
an orthonormal basis for the quantum Hilbert space, so the coefficient
vector~$x$ has indices~$n=0$, 1, \dots, $N$, where the number of
particles is limited to~$N$ to make the simulation finite.  (In the
experiment shown, $N=25$.)  The hamiltonian matrix is then diagonal,
with
\begin{equation}
	H_{nn} = \omega_n = n(n-9)/2.
\end{equation}
The eigenvalues range from~$\omega_4=\omega_5=-10$,
through~$\omega_1=\omega_9=0$, up to~$\omega_{25}=200$.

In terms of wave functions and Schr\"odinger's equation, this could
be regarded as the hamiltonian for a single particle, where~$x_n$
was the coefficient of a Lagrange polynomial~$p_n$ (not quite) in
the wave function.  It would be quite a wierd hamiltonian physically,
with terms in products of position and momentum.  On the other hand,
it makes physical sense in second quantisation, where it describes
a collection of particles occupying the same orbital, with mutual
repulsion.

The oscillator was initialised in a quantum-optical coherent state
\cite{glauber}, where the number of particles is Poisson distributed,
with
\begin{equation}\label{eq:vii}
	x_n = e^{-|\alpha|^2/2}\frac{\alpha^n}{\sqrt{n!}}.
\end{equation}
The experiment was run with~$\alpha=2$, which provides some small
initial coefficients with large eigenfrequencies, which are useful
for demonstrating the growth of the parasitic solutions.

\begin{figure}[htbp]
  \centering
  \label{fig:a}\includegraphics[width=\textwidth]{pev.eps}
  \caption{Growth of parasitic eigenvectors.  Dotted line shows the stability limit $\omega=2/h$, dashed line shows the predicted growth rate~$\frac{i}{h}\ln(\frac{h\omega}2)$.  (Tidy up fonts.)}
  \label{fig:i}
\end{figure}

The results of solving this Hamiltonian with the semi-implicit
method, with 4~iterations and time step~$h=0.03$, are shown in
\cref{fig:a}.  The hamiltonian matrix is diagonal, so the norms~$|x_n|$
are constant in the exact solution.  The right hand side of
\cref{eq:vii} is shown as the solid lines at~$t=0$ and~$t=0.09$,
and the values of~$|x_n|$ computed by the semi-implicit formula are
shown as the rulings between these lines.  The size of~$x_n$ is
plotted for all four iterations of the formula within each timestep.
\Cref{eq:viii} shows that, for~$\omega_n\gg 2/h$, the growth of~$x_n$
will still be geometric, resulting in straight lines on the logarithmic
scale of the graph.

Parasitic growth can be
seen at the left hand side, where~$\omega_n$ is large.  The limit
of stability is~$\omega_n=2/h\approx 67$, and the dotted line
indicates~$\omega_17=68$.  The dashed line shows the result of
growing~$|x_n|$ at the predicted rate~$\frac4h\ln(\frac{h\omega}2)$.
The undulations in the lines~$|x_n(t)|$ correspond to the growth
rates given by the different lines in \cref{fig:ii}, which apply
for the fraction of each step ending with a given iteration.

@@

The semi-implicit formula discussed here is only one of the formulae
that can be used to integrate ordinary differential equations, and
that could be applied to solve \cref{eq:iii} numerically.  However,
most of these formulae suffer from similar problems, and become
unstable when the eigenvalues of the hamiltonian matrix are large
compared to the time step.  The phenomenon of stiffness in discretised
partial differential equations is well known \cite{something}.

@@

The situation where stiff integration formulae would be appropriate
is when the physics being simulated occurs at slow timescales, but
the method by which the physical state is being represented
inadvertently allows the state to contain a component with a large
energy, that evolves on a fast timescale.  In the numerical example,
this occured due to the coefficient vector~$x$ including coefficients
with large~$n$, for which~$x_n$ was too small to have any effect
on the state.  As demonstrated, the formula caused these minute
components to grow quickly and dominate the actual solution.  It would
be possible to address this by using an integration formula that
was adapted to stiff differential equations.  But it might be better
to address this by changing the representation so that the simulated
state can not include such components, or to modify the hamiltonian
dynamics so that they evolved with reasonable frequencies in the
simulation.

In the example here, the representation was simply an expansion
over energy eigenstates.  This can be modified very simply, by
reducing~$N$ to exclude the large eigenvalues and frequencies.  In
the example, an expansion with coefficients~$x_0$ to~$x_{17}$ would
be stable with the time step used in the example.  However, this
is a contrived example.  If the energy eigenstates were known, their
amplitudes could be calculated directly as complex
exponentials~$e^{i\omega_nt}$, and there would be no need to use a
potentially unstable integration formula.

In real simulations, quantum states are represented by expansion
over a set of states that are not energy eigenstates, by use of a
variational ansatz, or by similar methods.  These representations
are not directly related to the hamiltonian, so it is often the
case that some states with spuriously high energies can be represented,
and the resulting ordinary differential equations are unnecessarily
stiff.  In these simulations, some care must be taken to avoid
inadvertently representing high energy states, or to modify the
hamiltonian dynamics to prevent them from causing stiffness.  Details
for one representation, using Gaussian wave packets, will come in
future papers.

@@

\bibliographystyle{siamplain}
\bibliography{references}
\end{document}

>>>>

% formula equations
The semi-implicit formula is a rule for integrating systems of
ordinary differential equations \cite{jkp-132-312,jkp-93-144}.  It
is most often used as a method of lines, to solve stochastic partial
differential equations that occur when phase space Monte-Carlo
methods are used in quantum optics and in atom optics \cite{pra-58-4824}.
These equations resemble Schr\"odinger's equation,
\begin{equation}\label{eq:vi}
    \frac{\partial\psi}{\partial t} = -i(-\nabla^2 + V)\psi,
\end{equation}
with additional nonlinear and random terms.  The spatial axes are
discretised, often using spectral derivative operators, to give a
system of ordinary or stochastic differential equations
\cite{2000-Trefethen-Spectral}.  Then these equations are integrated
by the semi-implicit formula.

@@

% basis coefficients
When the semi-implicit rule is applied to \cref{eq:vi}, or a
stochastic or nonlinear variant of it, the laplacian operator is usually
evaluated by a spectral method \cite{2000-Trefethen-Spectral}.  This is equivalent to expanding
the wave function~$\psi(x)$ over a basis of sinc wave packets
\begin{equation}
	\psi(x) = \sum_m s^m\phi_m(x),
\end{equation}
where, for a spatial grid of step~$l$,
\begin{equation}
	\phi_m(x) = \frac1{\sqrt{l}} \sinc\left(\textstyle\frac{x}{l} - m\right).
\end{equation}
This paper uses the convention
\begin{equation}
	\sinc x = \frac{\sin\pi x}{\pi x},
\end{equation}
under which the set of functions~$\phi_m$ is an orthonormal basis
for the space of functions limited to spatial bandwidth~$1/2l$.
The spectral method reduces Equation~\cref{eq:vi} to an ordinary
differential equation for the complex coefficient vector~$s$ of the
form
\begin{equation}\label{eq:iii}
	{ds\over dt}=-iHs,
\end{equation}
where~$H$ is a Hermitian matrix with elements
\begin{equation}
	H_{mn} = \int \phi_m(x) (-\nabla^2 + V(x)) \phi_n^\ast(x)\,dx.
\end{equation}

@@

Suppose that the point~$s_{n+1}$ also lies on this circle.
Then the projection of~$f_{n+\frac12}=-\frac{i}2H(s_n+s_{n+1})$ is parallel to the
chord from~$s_n$ to~$s_{n+1}$.  The midpoint of the chord can be
brought towards and away from the origin by moving~$s_{n+1}$ around
the circle, so there is some position for
which~$|s_{n+1}-s_n|=h|f_{n+\frac12}|$, so that~$s_{n+1}$
satisfies Equation~\cref{eq:i}.  The only error introduced by discretisation
is that~$c_k$ rotates with an angular frequency less
than~$\omega_k$, bounded by~$\pi/h$ as~$\omega_k$ becomes
large.

In practice, the iterations of~Equation~\cref{eq:vi} must be used
instead of~Equation~\cref{eq:i}, and this is less stable.  For
Hamiltonian dynamics with the form of Equation~\cref{eq:iii}, the
iteration rule becomes
\begin{equation}
s_{n+1}^{(j+1)}=s_n+hf_{n+\frac12}^{(j)}\qquad f_{n+\frac12}^{(j+1)}
	=-{i\over2}Hs_n-{i\over2}Hs_{n+1}^{(j)}
\end{equation}
or
\begin{equation}
s_{n+1}^{(j+1)}=\left(1-{ih\over2}H\right)s_n-{ih\over2}Hs_{n+1}^{(j)}
\end{equation}
with solution
\begin{equation}\label{eq:viii}
s_{n+1}^{(j)}=U^{(j)}s_n
	=\left(1+2\sum_{i=1}^j\left({-ihH\over 2}\right)^j\right)s_n.
\end{equation}
The time step matrix~$U^{(j)}$ for~$j$ iterations has the the same
eigenvectors~$v_k$ as the hamiltonian matrix~$H$.  Each of its
eigenvalues~$u_k^{(j)}$ is the sum of a geometric series
in~$-ih\omega_k/2$, which agrees to second order in~$h\omega_k$ with
the eigenvalue~$e^{-ih\omega_k}$ of the exact time step
operator~$U=e^{-ihH}$.  The relation between~$h\omega_k$ and~$u_k^{(j)}$ is
plotted in \cref{fig:ii}, for the first four iterations.

\begin{figure}[htbp]
  \centering
  \includegraphics[width=0.7\textwidth]{wev.eps}
  \caption{Relation between eigenvectors~$\omega_k$ of~$H$ and~$u_k^{(j)}$ of~$U^{(j)}$.  The lightest line is for~$j=1$, and the heavier ones are for~$j=2$, 3, 4.  The exact time evolution operator~$U$ has unimodular eigenvalues.}
  \label{fig:ii}
\end{figure}

The geometric series for~$U^{(j)}$ will converge if and only if the
angular frequencies~$\omega_k$ are bounded by the reciprocal time
step~$\frac2h$, a kind of Nyquist condition.  If the spectral
hamiltonian matrix has an eigenvalue~$\omega_k\gg \frac2h$, then~$U^{(j)}$
will enlarge the component of~$v_k$ in~$s_n$ by a factor close
to~$2(h\omega_k/2)^j$ over the time step~$h$.  This component is
thus a parasitic solution growing at the
rate~$\frac{\ln2}h+\frac{j}{h}\ln\left(\frac{h\omega_k}2\right)$.

The semi-implicit method is rarely used to solve the linear
Equation~\cref{eq:iii}, and more often used to solve a nonlinear
Gross-Pitaevskii~equation of the form
\begin{equation}
	{\partial\psi\over\partial t}=-i(\nabla^2+V+g|\psi|^2)\psi.
\end{equation}
However, following the usual approach to the stability of integation formulae \cite{1996-Trefethen-finite}, the mapping from~$\psi$ to $\partial\psi\over\partial t$ can be linearised at each time step, as
\begin{equation}
	\left.{\partial\psi\over\partial t}\right|_{\psi=\psi(t)+\delta\psi}
		\approx -i(\nabla^2+V+g|\psi(t)|^2)\psi(t) - i\hat H'(\psi(t))\delta\psi,
\end{equation}
where~$\hat H'(\psi(t))$ is an effective hamiltonian operator for
the field~$\delta\psi$.  This linearisation has been studied
extensively, and the dynamics generated by the hamiltonian~$\hat
H'$ are Bogoliubov sound waves \cite{aop-70-67}.  When this equation
is discretised with the semi-implicit formula, the stability condition
is that the Bogoliubov sound wave modes on a spatial grid of
spacing~$l$ have frequencies bounded by~$2/h$.  For a sufficiently
fine grid, the high-frequency Bogoliubov modes have the same
frequencies as the eigenvectors of~$H$ \cite{2002-Pethick-Bose}.
Most simulations use grids fine enough for this to be the case, and
the nonlinear term will not affect their stability.

@@

As a test of this analysis, a quantum Kerr oscillator was simulated
using the semi-implicit formula.  This system is an analog of the
fields that are usually simulated using the formula, where all the
particles are confined to the same orbital.  The states with definite
numbers of particles are an orthonormal basis for the quantum Hilbert
space, so the coefficient vector~$s$ has indices~$n=0$, 1, \dots,
$N$, where the possible number of particles is limited to~$N$.  In
the example run here, $N=25$.  The hamiltonian matrix is diagonal,
with
\begin{equation}
	H_{nn} = \omega_n = n(n-9)/2.
\end{equation}
The eigenvalues range from~$\omega_4=\omega_5=-10$,
through~$\omega_1=\omega_9=0$, up to~$\omega_{25}=200$.

The oscillator was initialised in a quantum optical coherent state
\cite{prx-131-2766} with a Poisson distribution of particle numbers,
\begin{equation}\label{eq:vii}
	s_n = e^{-|\alpha|^2/2}\frac{\alpha^n}{\sqrt{n!}}.
\end{equation}
The experiment was run with the coherent amplitude~$\alpha=2$.  This
provides some small initial coefficients with large eigenfrequencies,
which demonstrate the growth of the parasitic solutions.

\begin{figure}[htbp]
  \centering
  \label{fig:a}\includegraphics[width=\textwidth]{pev.eps}
  \caption{Growth of parasitic eigenvectors.  Dotted line shows the stability limit $\omega=2/h$, dashed line shows the predicted growth rate~$\frac{i}{h}\ln(\frac{h\omega}2)$.}
  \label{fig:i}
\end{figure}

The results of solving this Hamiltonian with the semi-implicit
method, with 4~iterations and time step~$h=0.03$, are shown in
\cref{fig:a}.  The hamiltonian matrix is diagonal, so the norms~$|s_n|$
are constant in the exact solution.  The right hand side of
Equation~\cref{eq:vii} is shown as the solid lines at~$t=0$ and~$t=0.09$,
and the values of~$|s_n|$ computed by the semi-implicit formula are
shown as the rulings between those lines.  The size~$|s_n|$ is
plotted for all four iterations of the formula within each timestep.
\Cref{eq:viii} shows that, for~$\omega_n\gg 2/h$, the growth of~$s_n$
from step to step is still geometric, resulting in straight lines on the logarithmic
scale of the graph.

{\bf Check growth rates.}

Parasitic growth can be
seen at the left hand side, where~$\omega_n$ is large.  The limit
of stability is~$\omega_n=2/h\approx 67$, and the dotted line
indicates~$n=17$, with~$\omega_{17}=68$.  The dashed line shows the result of
growing~$|s_n|$ at the predicted rate~$\frac4h\ln(\frac{h\omega}2)$.
The undulations in the lines~$|s_n(t)|$ correspond to the growth
rates given by the different lines in \cref{fig:ii}, each of which applies
for the initial fraction of each step up to the corresponding iteration.

@@

The semi-implicit formula discussed here is only one of the formulae
that can be used to integrate ordinary differential equations, and
that could be applied to solve Equation~\cref{eq:iii} numerically.  However,
most of these formulae suffer from similar problems, and become
unstable when the eigenvalues of the hamiltonian matrix are large
compared to the time step.  The phenomenon of stiffness in discretised
partial differential equations is well known \cite{something}.

@@

On the other hand, stiff integration formulae would be appropriate
where the dynamics being simulated is relatively slow, but the
representation of the quantum state inadvertently contains some
components with large energies whose phases oscillate rapidly.  This
was the case in the example shown above, where the coefficient
vector included coefficients with large~$n$, for which~$s_n$ was
too small to matter.  However, $\omega_n$ was large enough that the
formula caused these minute components to grow and dominate the
actual solution.  This could be addressed by using an integration
formula that was adapted to stiff differential equations.  But it
would be better to change the representation to exclude the high
energy components, or to modify the simulated dynamics so that these
components oscillate with reasonable frequencies.

The example above can be modified very simply, by reducing~$N$ to
exclude the high frequencies.  An expansion with coefficients~$s_0$
to~$s_{17}$ would be stable with the time step used above, and the
coefficients~$s_n$ with $n>17$ are small.  However, this is a
contrived example.  Since the energy eigenstates are known, their
amplitudes can be calculated directly as complex
exponentials~$e^{i\omega_nt}$, and there is no need to use any
integration formula, stable or otherwise.

In a real simulation, the energy eigenstates are not known beforehand,
and the quantum state is represented some other way.  This could
be an expansion over a set of states that are not energy eigenstates,
a variational ansatz, or a similar method.  Some of these representations
will allow states with spuriously high energies to be represented,
so that the differential equations satisfied by the representation
are unnecessarily stiff.  In this case, care must be taken to avoid
inadvertently representing high energy states, or to modify the
Hamiltonian dynamics to prevent them from causing stiffness.  Details
for one such representation, using Gaussian wave packets, will be
given in a future paper.

@@

\bibliographystyle{siamplain}
\bibliography{references,goats}
\end{document}
